Intro Outline

### 2020-06-08 doc/Intro.Rmd
small changes to v2

(1) Meiosis can be reduced to the expression of (2n -> 4n -> 2n -> 1n) which tracks the duplication of a diploid genome into haploid cell products. The meiotic program relies on crossovers and the process of recombination to ensure the correct separation of chromosomes. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). The gwRR regulates populations responses to selection, and determine the fate of novel mutations by transfering beneficial mutations onto novel genetic backgrounds or by breaking linkage of bad mutations from benefical backgrounds.

@@

(mention fitness more explicitly). The lower bound/threshold is thought to be an obligate crossover per homolog pair -- is a checkpoint for proper chromosome segregation (Hunt Nagaoka, Hassold). While too much recombination (ectopoic recombination) and DNA damage preceeding crossover formation (may be a constraint for an upper threshold. (cite other than Inoue, OttoPaysuer?).  **As the gwRR influences the evolution of other traits, it is important to document and understand the evolutionary patterns for the genome wide recombination rate itself.**

@@

1a. There is substantial in gwRR variation within species (across populations) and (across individuals)
•	While there are fewer measures from closely related species, variation at this level is more restricted to finer scales (hotspot level / recombination landscape level).

@@

•	Long history since the discovery -- in fruit flies and silk moths
Heterochiasmy was first documented ---

@@

As detailed in Sardell Kirkpatrick, (for a species with hetreochiasmy, rarely do they have a single recombination landscape. Instead there are broad scale differences recombination rate along

•	Most species have more recombination in females than males,

•	Crossover placement is sexually dimorphic: male telomeric, female uniform placement

•	SC length / meiotic chromosome length is longer in females of a few mammal species, but there are exceptions, Celegean (not much different) and A.thalnia (opposite direction) (Cahoon and Libdua).

@@

@@

>>>>

Meiosis can be reduced to the expression of (2n -> 4n -> 2n -> 1n) which tracks the duplication of a diploid genome into haploid cell products. The meiotic program relies on crossovers and the process of recombination to ensure the correct separation of chromosomes. The total number of COs per cell (at the 4n stage) is equal to the genome wide recombination rate (gwRR). The gwRR regulates populations responses to selection, and determine the fate of novel mutations by transfering beneficial mutations onto novel genetic backgrounds or by breaking linkage of bad mutations from benefical backgrounds.

@@

<(mention fitness more explicitly?)> The lower bound/threshold is thought to be an obligate crossover per homolog pair -- is a checkpoint for proper chromosome segregation (Hunt Nagaoka, Hassold). While too much recombination (ectopoic recombination) and DNA damage preceeding crossover formation (may be a constraint for an upper threshold. (cite other than Inoue, OttoPaysuer?).  **As the gwRR influences the evolution of other traits, it is important to document and understand the evolutionary patterns for the genome wide recombination rate itself.**

@@

1a. There is substantial in gwRR variation within species (across populations) and (across individuals)

•	While there are fewer measures from closely related species, variation at this level is more restricted to finer scales (hotspot level / recombination landscape level). **find a good citation for this**

@@

•	Long history since the discovery -- in fruit flies and silk moths at the turn of the 20th century (cite sturavant)
Heterochiasmy was first documented ---

@@

A early pattern of heterochiasmy that emerged was that in many species females have more recombination compared to males, considered female bias-heterochiasmy. However as observations have expanded to more species, more examples of male biased heterochiasmy have been observed; including cattle, marsupials, (fish)  (while birds have a smaller overall degree of heterochiasmy)

A more conserved feature of heterochiasmy As discussed in Sardell Kirkpatrick, are the broad scle differences in the placement of crossovers along chromosomes.

In male recombination landscapes, there is a bias towards the ends near the telomeres while in females crossovers are more uniformly placed (and occur closer to the centromere compared to males).

(the degree of this pattern varies across species (some have female biased telomere-bias, some have chromosome specific patterns and others have no significant sex difference. It's worth pointing out that the exceptions may require modifed assumptions as to how the meiotic program evolves; marsupials form chromosome chains during meiosis, some fish species has sequential hermaphrodism, and birds have many microchromosomes chromosomes. <grasshoppers?>

@@

<SC length / meiotic chromosome length is longer in females of a few mammal species, but there are exceptions, Celegean (not much different) and A.thalnia (opposite direction) (Cahoon and Libdua).>

@@

Overall house mouse -- using the house mouse model -- will better allow narrowing in on the causes // developing testable hypotheses for explaining any (particular patterns) variation discovered.
